---
title: "Data import into DHIS2"
author: "Jason Pickering"
date: "August 6, 2017"
output:
  slidy_presentation: default
  ioslides_presentation: default
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Set your login credentials and server

- You may need to change the following to suit your environment. 
- Set the `root.dir` variable to the GitHub directory for this project is. 
- Change `baseurl` to point to where your installation of DHIS2 is running. 
- Change `username` and `district` to suit your DHIS2 installation.


```{r , echo = TRUE, results='hold',message=FALSE}

knitr::opts_knit$set(root.dir= "/home/jason/development/dhis2-data-munging/")
baseurl<-"http://localhost:8080/dhis/"
username<-"admin"
password<-"district"

```


## Load some libraries and login to DHIS2

```{r , echo = FALSE, results='hold',message=FALSE}
require(httr)
require(jsonlite)
require(assertthat)
require(rlist)
require(reshape2)
require(plyr)
require(tidyjson)
require(listviewer)
generateUID <- function(codeSize = 11) {
  #Generate a random seed
  runif(1)
  allowedLetters <- c(LETTERS, letters)
  allowedChars <- c(LETTERS, letters, 0:9)
  #First character must be a letter according to the DHIS2 spec
  firstChar <- sample(allowedLetters, 1)
  otherChars <- sample(allowedChars, codeSize - 1)
  uid <- paste(c(firstChar,paste(
    otherChars, sep = "", collapse = "" )), sep = "", collapse = "")
  return(uid)
}
#This will give us a cookie we can use for later use. 

loginDHIS2<-function(baseurl,username,password) {
  url<-paste0(baseurl,"api/me")
  r<-GET(url,authenticate(username,password))
  assert_that(r$status_code == 200L) }

loginDHIS2(baseurl,username,password)


```


## Loading the data

- Get the data from the UNAIDS website
- It took some time to reverse engineer the API for this website, but use this as a guide. 
- We will assemble a request for the `http://aidsinfo.unaids.org/` site and parse this data into a data frame.


```{r}

body <- list("reqObj[TabStatus]" = "world", 
             "reqObj[Group_Name]" = "People living with HIV", 
             "reqObj[Display_Name]" = "Number of people living with HIV",
            "reqObj[TimePeriod]" = "2016",
            "reqObj[Area_Level]" = "2")
url<-"http://aidsinfo.unaids.org/datasheetdatarequest"
r <- POST(url, body = body, encode = "form")
d<-fromJSON(content(r,"text"))
r %>% content(.,"text") %>% jsonedit(mode = "code")

```


## Loading the data

- We have to spend a bit of time to look at the response format, but it seems to be a JSON object which we will need to reconstruct into a flat table. 
- Create two tables for the year and country
- Unnest the data contained within the array. 
- There are better ways to do this. :) 

```{r}

d_all<-data.frame()
#Loop through the nested data frames
for ( i in 1:length(d$tableData$Data_Val ) ) {
foo<-data.frame(d$tableData$Data_Val[[i]],stringsAsFactors = FALSE)
names(foo)<-as.vector(t(d$MultiSubgroups))
foo$country<-d$tableData$Area_Name[[i]]
foo$year<-d$tableYear
d_all<-rbind(d_all,foo)
}

knitr::kable(head(d_all))
```


## More data cleanup

- Lets only get the middle estimates, ignoring the lower and upper estimates. 
- Lets round these numbers to the nearest integer.
- Remove any "NA" or **NULL** values. They will not be imported.

```{r echo=TRUE, messages=FALSE}
d<-d_all[,c("All ages estimate","country","year")]
names(d)<-c("value","orgUnit","period")
d<-d[!is.na(d$value),]
d$value<-round(as.numeric(d$value))
knitr::kable(head(d))
```



## Matching the orgunits from DHIS2


- Get a listing of countries from DHIS2. We will need to match these with our data. 
- Convert the country names in the data file back to a character (instead of a factor) so that we can use this column to join the OU listing from DHIS2 with it. 

```{r}
r <- GET(paste0(baseurl,"api/27/organisationUnits?paging=false&filter=level:eq:3&fields=id,name"))
r<- httr::content(r, "text")
ous<-jsonlite::fromJSON(r,flatten=TRUE)$organisationUnits
names(ous)<-c("ou_name","ou_id")
d$orgUnit<-as.character(d$orgUnit)
#Now, we need to try and merge the OUs from DHIS2 with the OUs from the CSV file. 
#This will be similar to a "left join" as we will keep all of the 
```


- Very likely, this is not going to match, so let's try to first perform a **LEFT JOIN** on the data. 
- This will keep all of the countries, even if there is not a match from DHIS2. 


```{r}

d_merged<-merge(d,ous,by.x="orgUnit",by.y="ou_name",all.x=TRUE)
unique(d_merged[is.na(d_merged$ou_id),"orgUnit"])
```


- So, we see we have a number of countries which do not match. 
- We can define a mapping to convert the names in the data file to the names in DHIS2.
- We are still going to be missing a few countries (like South Sudan ) but our hierarchy does not have this country, so we will just have to ignore it. 


```{r}
from<-c("Bolivia (Plurinational State of)",
"Cabo Verde","Côte d'Ivoire","Democratic People's Republic of Korea",
"Myanmar","Republic of Korea",
"Russian Federation","South Sudan",
"Venezuela (Bolivarian Republic of)")
to<-c("Bolivia","Cape Verde","Cote d'Ivoire","Korea, Democratic People's Republic of",
      "Burma","Korea, Republic of","Russia","South Sudan","Venezuela")
knitr::kable(data.frame(from,to))
```

- The mapping looks OK, so lets reassign the names.
- This time we will perform an **INNER JOIN** and discard any data which does not have a matching country in DHIS2. 

```{r}
d$orgUnit<-plyr::mapvalues(d$orgUnit,from,to,warn_missing = FALSE)
d_merged<-merge(d,ous,by.x="orgUnit",by.y="ou_name")
d_merged<-d_merged[,c("ou_id","period","value")]
names(d_merged)<-c("orgUnit","period","value")
```




## Create a new data element for the data

- We will need to create a new data element for the data in the system.

```{r}
de_name<-"NUMBER OF PEOPLE LIVING WITH HIV"
set.seed(587838)
this_id<-generateUID()
des_import<-data.frame(name=de_name
                       ,shortName=de_name,
                       aggregationType="SUM",
                       valueType="NUMBER",
                       domainType="AGGREGATE",
                       id=this_id)

url<-paste0(baseurl,"api/27/metadata?importStrategy=CREATE")
#Post to the metadata API as JSON
r<-POST(url,body=toJSON(list(dataElements = des_import),
                        auto_unbox = TRUE),
        content_type_json())
```


- We get the data prepared to import into DHIS2. 
- Set all of the property names according to those used in the documentation. 
- Upload to DHIS2 and hope that it works.
- Finally, trigger analytics.

```{r}

d_merged$dataElement=this_id
d_merged<-plyr::colwise(as.character)(d_merged)

url<-paste0(baseurl,"api/27/dataValueSets?preheatCache=true")
r<-POST(url,body=toJSON(list(dataValues = d_merged),auto_unbox = TRUE),content_type_json())
#Lets trigger analytics
url<-paste0(baseurl,"api/27/resourceTables/analytics")
r<-POST(url)

```